Is Google Translate Sexist? -- Assessing Gender Bias in Machine Translation

### 2017-12-17 paper/main.tex
Updated paper

\title{Assessing Gender Bias in Machine Translation}
\author{Marcelo Prates \and Luis Lamb}

@@

\begin{abstract}
\end{abstract}

@@

Chomsky and Norvig's debate \cite{norvig2017chomsky} is a microcosmos of the two leading standpoints about the future of science in the face of increasingly sophisticated statistical models. Are we, as Chomsky seems to argue, jeopardizing science by relying on statistical tools to perform predictions instead of perfecting traditional science models, or are these tools, as Norvig argues, components of the scientific standard since its conception? Currently there are no satisfactory resolutions to this conundrum, but perhaps statistical models pose an even greater and more urgent threat to our society. On a 2014 article, Londa Schiebinger suggested that scientific research fails to take gender issues into account, arguing that the phenomenon of male defaults on new technologies such as Google Translate provides a window into this asymmetry \cite{schiebinger2014scientific}. Since then, recent worrisome results in machine learning have somewhat supported Schiebinger's view, and perhaps even partially confirmed some of Chomsky's fears. Not only Google photos' statistical image labeling algorithm has been found to classify dark-skinned people as gorillas \cite{garcia2016racist} and purportedly intelligent programs have been suggested to be negatively biased against black prisoners when predicting criminal behavior \cite{angwin2016machine} but the machine learning revolution has also indirectly revived heated debates about the controversial field of physiognomy, with proposals of AI systems capable of identifying the sexual orientation of an individual through its facial characteristics \cite{wang2017deep}. \emph{Machine bias}, the phenomenon by which trained statistical models unbeknownst to their creators become vessels of their own prejudices, is growing into a pressing concern for the modern times, and is an invitation for us to ask ourselves whether there are limits to our dependence on these techniques -- and more importantly, whether some of these limits have already been traversed.

With this in mind, we propose a quantitative analysis of the phenomenon of gender bias in machine translation. We believe this can be done by simply exploiting Google Translate to map sentences from a gender neutral language to English. As Figure \ref{fig:screenshot-gtranslate-hungarian} exemplifies, this approach

@@

@@

There is a prohibitively large class of nouns and adjectives that could in principle be substituted in the templates of Table \ref{tab:templates}. To simplify our dataset, we have decided to obtain a comprehensive list of professional occupations, which, we believe, are an interesting window into the nature of gender bias. Once again we resorted to a Wikipedia article (\url{https://en.wikipedia.org/wiki/Lists_of_occupations}) to collect this data, of which the statistics of occupations per category (Artistic, Corporate, Theatre, etc.) are shown in Table \ref{tab:occupations}. Finally, Table \ref{tab:occupations-examples} shows thirty examples of randomly selected occupations from our dataset.

@@

@@

While grouping translations by category helps shed light on the stereotypical gender roles among different professions, grouping translations by language can help us understand the effect each different culture has on this issue. Table \ref{tab:gender-by-language} for instance shows some sex ratios even larger than the previous ones, particularly when translating from Yoruba ($25.333$), Chinese ($27.5$) and Japanese, the last one peaking at an impressive ratio of $107.5$ male per female pronouns. Figure \ref{fig:gender-by-language} shows that, when grouping by language, gender neutral pronouns can be more prominent than male pronouns at least in one case: translating sentences from Basque yields 153 neutral vs 50 male and 5 female pronouns. Unfortunately this is the exception rather than the rule, with Yoruba following after with 131 neutral vs 304 male and 12 female pronouns. In total, female pronouns add up to $18.687\%$ among all categories, while male pronouns add up to $81.626\%$ and gender neutral pronouns to $8.466\%$, yielding an average sex ratio of $4.368$ ($14.405\%$ larger than what we get from grouping among categories). It should be noted however that Japanese and Basque, the two languages which stood out from the behavior observed in Figure \ref{fig:gender-by-category}, are precisely the two that Google Translate found hardest to translate. These findings should, as a result, be taken with a grain of salt.

@@

@@

\bibliographystyle{unsrt}
\bibliography{main}

>>>>

\title{Is Google Translate Sexist? -- Assessing Gender Bias in Machine Translation}
\author{Marcelo Prates \and Luis Lamb}

@@

\begin{abstract}
In recent times there has been a growing concern both inside and outside academia about the phenomenon of machine bias, where trained statistical models unbeknownst to their creators become vessels of their own prejudices. A large number of AI tools have recently been suggested to be harmfully biased towards some minority, with reports of racist criminal behavior predictors, Apple's Iphone X failing to differentiate between two Asian people and the infamous case of Google photos' classifying black people as gorillas. Although a systematic study of such biases can be difficult, we believe that automated translation tools can be exploited through gender neutral languages to yield a window into the phenomenon of \emph{gender} bias in AI. In this paper, we construct sentences with gender neutral pronouns in different languages which support them and translate these sentences into English using the Google Translate API. The resulting English sentences are constructed either with male, female or neutral pronouns, which allows us to collect statistics about the frequency of male defaults and evaluate which factors can influence the gender Google Translate assumes for each sentence. We show that not only there is a tendency towards translating pronouns to \say{he} but also that this tendency is exaggerated for sentences about male dominated fields such as computer science. At the other hand, sentences about artistic fields are less likely to yield male defaults. This possibly reflects the implicit assumptions our society has about stereotypical gender roles, which suggests that without proper care statistical machine translation can become hostage to the prejudices of those people behind its training dataset -- ourselves.
\end{abstract}

@@

Chomsky and Norvig's debate \cite{norvig2017chomsky} is a microcosmos of the two leading standpoints about the future of science in the face of increasingly sophisticated statistical models. Are we, as Chomsky seems to argue, jeopardizing science by relying on statistical tools to perform predictions instead of perfecting traditional science models, or are these tools, as Norvig argues, components of the scientific standard since its conception? Currently there are no satisfactory resolutions to this conundrum, but perhaps statistical models pose an even greater and more urgent threat to our society. On a 2014 article, Londa Schiebinger suggested that scientific research fails to take gender issues into account, arguing that the phenomenon of male defaults on new technologies such as Google Translate provides a window into this asymmetry \cite{schiebinger2014scientific}. Since then, recent worrisome results in machine learning have somewhat supported Schiebinger's view, and perhaps even partially confirmed some of Chomsky's fears. Not only Google photos' statistical image labeling algorithm has been found to classify dark-skinned people as gorillas \cite{garcia2016racist} and purportedly intelligent programs have been suggested to be negatively biased against black prisoners when predicting criminal behavior \cite{angwin2016machine} but the machine learning revolution has also indirectly revived heated debates about the controversial field of physiognomy, with proposals of AI systems capable of identifying the sexual orientation of an individual through its facial characteristics \cite{wang2017deep}. Similar concerns are growing at an unprecedented rate in the media, with reports of Apple's Iphone X face unlock feature failing to differentiate between two different Asian people \cite{womanunlockphone2017} and automatic soap dispensers which reportedly do not recognize black hands \cite{racistsoapdispenser2017}. \emph{Machine bias}, the phenomenon by which trained statistical models unbeknownst to their creators become vessels of their own prejudices, is growing into a pressing concern for the modern times, and is an invitation for us to ask ourselves whether there are limits to our dependence on these techniques -- and more importantly, whether some of these limits have already been traversed.

With this in mind, we propose a quantitative analysis of the phenomenon of gender bias in machine translation. We believe this can be done by simply exploiting Google Translate to map sentences from a gender neutral language to English. As Figure \ref{fig:screenshot-gtranslate-hungarian} exemplifies, this approach produces results consistent with the hypothesis that senteces about stereotypical gender roles are translated accordingly with high probability: \emph{nurse} and \emph{baker} are translated with female pronouns while \emph{engineer} and \emph{CEO} are translated with male ones.

@@

If automatic translation tools are indeed gender biased as Figure \ref{fig:screenshot-gtranslate-hungarian} seems to suggest, then possibly this phenomenon can be probed to yield many valuable insights. One could in principle group different professional occupations according to their broad category (artistic, scientific, industrial, etc.) and evaluate whether the proportion of male pronouns is more pronounced in some fields compared to others. We expect, for example, that most sentences about computer-related jobs will be translated with male pronouns, as a reflection of the gender asymmetry of this job market. We can also collect translation statistics among varied gender neutral languages to see whether different cultures make different assumptions about the gender of a subject.

If conclusive, in our view such an analysis would suggest that automatic translation tools should be designed in such a way as to avoid becoming hostages to their own underlying statistical models. Because these tools are trained with real world data they implicitly absorb the biases and stereotypes of our society, which must then be painstakingly removed. This poses an ethical rather than a technical limitation to statistical machine translation and big data statistical models in general.

@@

There is a prohibitively large class of nouns and adjectives that could in principle be substituted in the templates of Table \ref{tab:templates}. To simplify our dataset, we have decided to obtain a comprehensive list of professional occupations, which, we believe, are an interesting window into the nature of gender bias. Once again we resorted to a Wikipedia article (\url{https://en.wikipedia.org/wiki/Lists_of_occupations}) to collect this data, of which the statistics of occupations per category (Artistic, Corporate, Theatre, etc.) are shown in Table \ref{tab:occupations}. Finally, Table \ref{tab:occupations-examples} shows thirty examples of randomly selected occupations from our dataset. Finally, we have selected a small list of 21 adjectives, presented in Table \ref{tab:adjectives}.

@@

\begin{table}[H]
	\centering
	\begin{tabular}{|c|c|c|}
	\hline
	happy 		& sad 		& right 	\\ \hline
	wrong 		& afraid	& brave 	\\ \hline
	smart		& dumb		& proud 	\\ \hline
	ashamed		& kind		& cruel 	\\ \hline
	envious		& loving	& hateful 	\\ \hline
	modest 		& arrogant	& guilty	\\ \hline
	innocent	& helpless	& shy		\\ \hline
	\end{tabular}
	\caption{Adjectives}
	\label{tab:adjectives}
\end{table}

@@

While grouping translations by category helps shed light on the stereotypical gender roles among different professions, grouping translations by language can help us understand the effect each culture possibly has on this issue. Table \ref{tab:gender-by-language} shows some sex ratios even larger than the previous ones, particularly when translating from Yoruba ($25.333$), Chinese ($27.5$) and Japanese, the last one peaking at an impressive ratio of $107.5$ male per female pronouns. Figure \ref{fig:gender-by-language} shows that, when grouping by language, gender neutral pronouns can be more prominent than male pronouns at least in one case: translating sentences from Basque yields 153 neutral vs 50 male and 5 female pronouns. Unfortunately this is the exception rather than the rule, with Yoruba following after with 131 neutral vs 304 male and 12 female pronouns. In total, female pronouns add up to $18.687\%$ among all categories, while male pronouns add up to $81.626\%$ and gender neutral pronouns to $8.466\%$, yielding an average sex ratio of $4.368$ ($14.405\%$ larger than what we get from grouping among categories). It should be noted however that Japanese and Basque, the two languages which stood out from the behavior observed in Figure \ref{fig:gender-by-category}, are precisely the two that Google Translate found hardest to translate. These findings should, as a result, be taken with a grain of salt.

@@

Figures \ref{fig:histogram-female}, \ref{fig:histogram-female} and \ref{fig:histogram-neutral} shed further light on the asymmetry between the distribution male and female pronouns. While the number of occurences of male pronouns tends towards a normal distribution, the figure is changes drastically for the opposite gender, whose pronoun distribution is apparently governed by an inverse correlation. The behavior repeats itself for the gender neutral pronouns in Figure \ref{fig:histogram-neutral}.

\begin{figure}[H]
	\centering
	\includegraphics[width=\linewidth]{pictures/histogram-female}
	\caption{Histogram of the distribution of female pronouns among different occupation categories, with a distinctive peak at 0. The corresponding histogram for male pronouns (see Figure \ref{fig:histogram-male}), by constrast, peaks at 9.}
	\label{fig:histogram-female}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\linewidth]{pictures/histogram-male}
	\caption{Histogram of the distribution of male pronouns among different occupation categories, with a distinctive peak at 9. In contrast with the corresponding histograms for female and gender neutral pronouns (see Figures \ref{fig:histogram-female} and \ref{fig:histogram-neutral} respectively) which exhibit inverse correlations, here one can see a tendency towards normally-distributed pronoun counts.}
	\label{fig:histogram-male}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\linewidth]{pictures/histogram-neutral}
	\caption{Following the behavior observed for female pronouns, the histogram for the distribution of gender neutral pronouns among different categories shows a tendency towards an inverse correlation, with a distinctive peak at 0.}
	\label{fig:histogram-neutral}
\end{figure}

Finally, Figure \ref{fig:barplot-adjectives} and Table \ref{tab:gender-by-adjective} show how stereotypical gender roles possibly play a part when simple adjectives are translated. We observed that objective statements such as \say{he/she is wrong/right/guilty/innocent} are biased towards male defaults, while statements concerning emotional states (\say{sad},\say{kind},\say{shy}) amass at the opposite extreme of the sex ratio spectrum. Unsurprisingly, the statement \say{he/she is attractive} is translated predominately with female pronouns.

\begin{table}[H]
	\centering
	\begin{tabular}{|c|c|c|c|c|c|}
	\hline
	Adjective 	& Female & Male & Neutral & Ratio & Total \\ \hline
	\hline
	Shy     	& 6  & 2   & 2  & 0.333 	& 12  \\ \hline
	Attractive 	& 4  & 2   & 4  & 0.500   	& 12  \\ \hline
	Happy   	& 5  & 3   & 2  & 0.600   	& 12  \\ \hline
	Kind    	& 4  & 3   & 1  & 0.750 	& 12  \\ \hline
	Ashamed 	& 4  & 5   & 1  & 1.250 	& 12  \\ \hline
	Smart   	& 2  & 5   & 3  & 2.500   	& 12  \\ \hline
	Envious 	& 2  & 6   & 1  & 3.000   	& 12  \\ \hline
	Sad     	& 2  & 6   & 2  & 3.000   	& 12  \\ \hline
	Loving  	& 2  & 6   & 2  & 3.000   	& 12  \\ \hline
	Helpless 	& 2  & 6   & 2  & 3.000   	& 12  \\ \hline
	Brave   	& 2  & 7   & 1  & 3.500   	& 12  \\ \hline
	Proud   	& 2  & 7   & 1  & 3.500   	& 12  \\ \hline
	Hateful 	& 1  & 5   & 3  & 5.000   	& 12  \\ \hline
	Dumb    	& 1  & 6   & 2  & 6.000   	& 12  \\ \hline
	Innocent 	& 1  & 8   & 1  & 8.000   	& 12  \\ \hline
	Right   	& 0  & 7   & 3  & -   		& 12  \\ \hline
	\hline
	Total   	& 43 & 129 & 41 & 3     	& 264 \\ \hline
	\end{tabular}
	\caption{Number of female, male and neutral pronominal genders in the translated sentences for each selected adjective.}
	\label{tab:gender-by-adjective}
\end{table}

\begin{figure}[H]
	\centering
	\includegraphics[width=\linewidth]{pictures/barplot-adjectives}
	\caption{The distribution of pronominal genders for each word in Table \ref{tab:adjectives} shows how stereotypical gender roles can play a part on the automatic translation of simple adjectives. One can see that adjectives such as \emph{shy} and \emph{attractive} are predominantly translated with female pronouns, while words like \emph{guilty}, \emph{innocent}, \emph{wrong}, \emph{right}, \emph{arrogant} are almost exclusively translated with male pronouns. Objective statements have a tendency towards male defaults, while statements concerning emotional states (\emph{shy}, \emph{happy}, \emph{kind}, \emph{ashamed}) amass at the other extreme of the sex ratio spectrum.}
	\label{fig:barplot-adjectives}
\end{figure}

@@

In this paper, we have provided evidence that statistical translation tools such as Google Translate exhibit implicit gender biases and a tendency towards male defaults, which possibly stem from the real world data used to train them. As a result, we suggest that such tools can be probed to yield insights about stereotypical gender roles in our society. By translating sentences from gender neutral languages such as Hungarian and Chinese into English, we were able to collect statistics about the asymmetry between female and male pronominal genders in the translation process. Because Google Translate typically uses English as a \emph{lingua franca} to translate between other languages, our findings probably generalize to most translations from gender neutral idioms, although we have not tested this hypothesis. Our results further show that although male pronouns are typical, the proportion of female pronouns can vary significantly according either to the adjectives used (\say{shy}/\say{happy} vs \say{brave}/\say{Guilty}) or the category of professional occupations (artistic jobs are far more likely to be translated with female pronouns than computer-related ones). We have also shown that different languages are differently biased towards male defaults, with Hungarian exhibiting a better balance between male and female pronouns than, say, Chinese. Some languages such as Yoruba and Basque have been found to translate sentences with gender neutral pronouns very often, although this is the exception rather than the rule.

We think this work can shed further light on some of the technical and ethical difficulties that arise from statistical machine translation, and hope that it arouses discussions about the role of AI engineers on minimizing the harmful effects of the pressing modern concern of machine bias.

\clearpage
\bibliographystyle{unsrt}
\bibliography{main}
